---
title: "BEAM walkthrough - grey seal, harbour seal and combined pinniped case"
author: "WGBYC ToR C group"
date: "2026-09-24"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE
)
```

# Purpose

This walkthrough adds one pooled pinniped case to the standard BEAM workflow while leaving BEAM modelling and raising unchanged.

The four fitted cases are:

- `halichoerus grypus` - grey seal;
- `phoca vitulina` - harbour seal;
- `grey_harbour` - grey + harbour + unresolved `phocidae`/`pinnipedia`;
- `grey_harbour_identified` - grey + harbour only, used as a diagnostic.

The flow is deliberately simple:

`fetch -> standard BEAM cleaning -> build pinniped cases -> calc_bpue() -> calc_total() -> compare -> plot`

The pooled cases are constructed before `calc_bpue()`. After that, the standard BEAM functions are used unchanged. Composite ICES area codes containing `~` are not split.

# 1. Setup

## 1.1 Packages

```{r libraries}
library(data.table)
library(reshape2)
library(icesConnect)
library(httr)
library(jsonlite)
library(glmmTMB)
library(metafor)
library(ggeffects)
library(emmeans)
library(this.path)
library(doSNOW)
library(doParallel)
library(progress)
library(stringdist)
library(stringi)
library(taxizedb)
library(countrycode)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(scales)
library(sf)
library(rnaturalearth)

setwd(this.path::this.dir())
source("lib/beam_lib.R")
source("lib/pinniped_case.R")
```

## 1.2 Controls

These are the settings most likely to change between runs.

```{r controls}
years <- 2017:2025
final_year <- max(years)

# Refresh the standard BEAM D1/D2/D3 cache when required.
FORCE_DOWNLOAD <- FALSE

# Parallel processing is optional.
USE_PARALLEL <- TRUE
N_CORES <- 12

# Area x métier figures. The case-comparison plot uses the same combinations.
N_STANDOUT_AREA_METIER <- 10
SHOW_DIAGNOSTIC_CASE <- FALSE

# Map choice: "top_n" or "all".
MAP_SHOW <- "top_n"
N_MAP_AREAS <- 10

stopifnot(MAP_SHOW %in% c("top_n", "all"))

# ICES primary colour palette.
ICES_BLUE <- "#002B5F"
ICES_GREEN <- "#649388"
ICES_HORIZON <- "#3499CC"
ICES_ORANGE <- "#E4612B"
ICES_SEASHELL <- "#F4F0E8"

CASE_COLOURS <- c(
  "Grey seal" = ICES_GREEN,
  "Harbour seal" = ICES_HORIZON,
  "Grey + harbour (identified only; diagnostic)" = ICES_BLUE,
  "Grey + harbour + unresolved pinnipeds" = ICES_ORANGE
)
```

Figures use the ICES primary colour palette: ICES blue, ICES green, Horizon blue, Rescue orange and Seashell white.

```{r credentials}
ices_username <- readLines("username.txt", warn = FALSE)
ices_token <- readLines("token.txt", warn = FALSE)
```

Optional parallel backend. If you do not need it, set `USE_PARALLEL <- FALSE` above.

```{r cluster}
# Optional warp speed, pun intended, for the model-fitting step.
if (USE_PARALLEL) {
  ncores <- min(N_CORES, parallel::detectCores())
  cl <- parallel::makeCluster(ncores)
  registerDoSNOW(cl)
}
```

# 2. Fetch and clean the standard BEAM data

```{r get-data}
all1 <- beam_get_raw_data(
  "data/all1.csv",
  years = years,
  force_download = FORCE_DOWNLOAD
)

obs1 <- beam_get_raw_data(
  "data/obs1.csv",
  years = years,
  force_download = FORCE_DOWNLOAD
)

bycatch1 <- beam_get_raw_data(
  "data/bycatch1.csv",
  years = years,
  force_download = FORCE_DOWNLOAD
)
```

Keep the annual corrections aligned with the current BEAM master walkthrough. The block below contains the 2025 GB and Portugal corrections used for this assessment.

```{r annual-fixes}
# GB VL0010 2025 correction.
vl0010_2025 <- subset(
  all1,
  vesselLengthRange == "VL0010" & year == 2024 & country == "GB"
)
vl0010_2025$year <- 2025
all1 <- rbind(all1, vl0010_2025)

obs1$country[obs1$country == "GB-NIR"] <- "GB"

# Portugal GNS/GTR fishing-effort correction.
sel <- all1[
  ecoregion == "Bay of Biscay and the Iberian Coast" &
    metierL4 == "GNS" &
    country == "PT"
]

new <- rbind(
  copy(sel)[, `:=`(metierL4 = "GTR", daysAtSeaF = daysAtSeaF * 0.69)],
  copy(sel)[, daysAtSeaF := daysAtSeaF * 0.31]
)

all1 <- rbind(
  all1[!(ecoregion == "Bay of Biscay and the Iberian Coast" &
           metierL4 == "GNS" &
           country == "PT")],
  new
)

fwrite(
  all1,
  "data/pinniped_case_all1_postcorrection.csv",
  sep = ";",
  na = "NA"
)
```

Run the normal BEAM preparation scripts.

```{r standard-cleaning}
source("lib/clean_data.R")
source("lib/generate_the_list.R")

obs3_standard <- copy(obs3)
bycatch3_standard <- copy(bycatch3)
```

# 3. Build the pinniped cases

The helper does the only case-specific data preparation: it creates pooled monitoring/bycatch inputs without double-counting the monitoring denominator.

```{r prepare-pinnipeds}
pinniped <- prepare_pinniped_case(
  obs3_standard = obs3_standard,
  bycatch3_standard = bycatch3_standard,
  obs2 = obs2,
  bycatch2 = bycatch2,
  all2 = all2,
  results_dir = "results"
)

obs3_pinniped <- pinniped$obs3
all2_pinniped <- pinniped$all2

# Standard downstream BEAM functions use the global obs3 object.
obs3 <- obs3_pinniped
```

## 3.1 Quick audit

Unresolved pinnipeds are already included in `grey_harbour`; these tables simply make them visible.

```{r unresolved}
pinniped$unresolved_totals
pinniped$unresolved_area_summary
```

The detailed audits are written to `results/pinniped_case_unclassified_*.csv`.

These are the rows you are looking for: the denominator/matching QC tables below should normally be empty.

```{r prep-qc}
pinniped$count_qc
pinniped$monitoring_denominator_qc
pinniped$unmatched_pooled_bycatch
pinniped$unmatched_identified_pool_bycatch
```

# 4. Estimate BPUE with standard BEAM

From here, the normal BEAM model-fitting functions are used.

```{r bpue}
area_cols <- c("ecoregion", "areacode", "metierl4", "species")
area_m4_spec <- unique(obs3_pinniped[, ..area_cols])

bpue1_pinniped_area <- calc_bpue(
  needle = area_m4_spec,
  cols = area_cols,
  dat = obs3_pinniped[taxon_bycatch_monitor_ok == TRUE],
  years = years,
  include.weights = TRUE
)

fwrite(
  bpue1_pinniped_area,
  "data/pinniped_case_bpue1_area.csv",
  sep = ";",
  na = "NA"
)
```

Each exact `ecoregion x areacode x métier L4 x case` combination is passed separately to `calc_bpue()`.

# 5. Annotate BPUE

```{r annotation}
bpue2_pinniped_area <- annotate_bpue(
  bpue1_pinniped_area,
  cols = c("ecoregion", "areacode", "metierl4")
)

bpue2_pinniped_area <- add_pinniped_diagnostics(
  bpue = bpue2_pinniped_area,
  obs3_pinniped = obs3_pinniped,
  cols = area_cols
)
```

The standard BEAM zero/NA handling is left untouched. This compact diagnostic only describes what came back.

```{r bpue-diagnostics}
bpue2_pinniped_area[
  result_status != "BPUE estimated",
  .(
    ecoregion, areacode, metierl4, species,
    observed_n, monitored_days, n_monitoring_cells,
    model, bpue, result_status
  )
]
```

# 6. Raise using final-year fishing effort

`clean_data.R` reduces D1 to the final assessment year, so for this run the raising effort is `r final_year` fishing effort.

```{r total}
stopifnot(all(unique(all2_pinniped$year) == final_year))

tot1_pinniped_area <- calc_total(
  bpue = bpue1_pinniped_area,
  cols = area_cols,
  obs = obs3_pinniped[taxon_bycatch_monitor_ok == TRUE],
  allx = all2_pinniped,
  years = years,
  include.weights = TRUE,
  verbose = FALSE
)

fwrite(
  tot1_pinniped_area,
  "data/pinniped_case_tot1_area.csv",
  sep = ";",
  na = "NA"
)
```

# 7. Final results

```{r final-results}
total_cols <- intersect(
  c(area_cols, "tot_mean", "tot_lwr", "tot_upr", "message", "fishing_effort"),
  names(tot1_pinniped_area)
)

pinniped_results <- merge(
  bpue2_pinniped_area,
  tot1_pinniped_area[, ..total_cols],
  by = area_cols,
  all.x = TRUE
)

unresolved_final <- pinniped$unresolved_area_summary[
  year == final_year,
  .(unclassified_observed_final = sum(n_unclassified, na.rm = TRUE)),
  by = .(ecoregion, areacode, metierl4)
]

pinniped_results[
  unresolved_final,
  on = .(ecoregion, areacode, metierl4),
  unclassified_observed_final := i.unclassified_observed_final
]

pinniped_results[
  is.na(unclassified_observed_final),
  unclassified_observed_final := 0
]

pinniped_results[
  species != "grey_harbour",
  unclassified_observed_final := NA_real_
]

pinniped_results[, case_label := fcase(
  species == "halichoerus grypus", "Grey seal",
  species == "phoca vitulina", "Harbour seal",
  species == "grey_harbour_identified", "Grey + harbour (identified only; diagnostic)",
  species == "grey_harbour", "Grey + harbour + unresolved pinnipeds",
  default = species
)]

pinniped_results[, case_role := fcase(
  species == "grey_harbour_identified", "diagnostic",
  default = "reporting"
)]

pinniped_results <- add_pinniped_total_status(pinniped_results)

setcolorder(
  pinniped_results,
  c(
    "case_label", "case_role", "species", "ecoregion", "areacode", "metierl4",
    setdiff(
      names(pinniped_results),
      c("case_label", "case_role", "species", "ecoregion", "areacode", "metierl4")
    )
  )
)

fwrite(
  pinniped_results,
  sprintf("results/pinniped_case_area_results_%s.csv", final_year),
  sep = ";",
  na = "NA"
)
```

Rows without a raised estimate can be reviewed here:

```{r final-diagnostics}
pinniped_results[
  total_status != "raised estimate available",
  .(
    case_label, ecoregion, areacode, metierl4,
    observed_n, monitored_days, model, bpue,
    result_status, total_status, message
  )
]
```

# 8. Compare combined and species-specific estimates

The main comparison is at `ICES area x métier L4`.

A missing grey or harbour estimate is treated as zero only when that case has zero observed animals in the stratum. Positive observed bycatch with no raised estimate stays `NA`.

The key quantities are:

- `raised_difference` = full combined - (grey + harbour);
- `percent_increase` = `raised_difference / (grey + harbour) x 100`;
- `pooling_effect` = identified-only pooled - (grey + harbour);
- `unresolved_increment` = full combined - identified-only pooled.

The percentage is undefined when the grey + harbour denominator is zero.

```{r combined-comparison}
# One comparison table to bind the cases together.
case_extract <- function(dat, species_name, prefix) {
  dat %>%
    filter(species == species_name) %>%
    select(
      ecoregion, areacode, metierl4,
      observed_n, tot_mean, tot_lwr, tot_upr, total_status
    ) %>%
    rename_with(
      ~ paste0(prefix, "_", .x),
      c(observed_n, tot_mean, tot_lwr, tot_upr, total_status)
    )
}

combined_full <- case_extract(pinniped_results, "grey_harbour", "combined") %>%
  filter(!is.na(combined_tot_mean))

grey_case <- case_extract(pinniped_results, "halichoerus grypus", "grey")
harbour_case <- case_extract(pinniped_results, "phoca vitulina", "harbour")
identified_pool <- case_extract(
  pinniped_results,
  "grey_harbour_identified",
  "identified_pool"
)

comparison_area_metier <- combined_full %>%
  left_join(grey_case, by = c("ecoregion", "areacode", "metierl4")) %>%
  left_join(harbour_case, by = c("ecoregion", "areacode", "metierl4")) %>%
  left_join(identified_pool, by = c("ecoregion", "areacode", "metierl4")) %>%
  mutate(
    grey_zero_assumed = is.na(grey_tot_mean) & coalesce(grey_observed_n, 0) == 0,
    harbour_zero_assumed = is.na(harbour_tot_mean) & coalesce(harbour_observed_n, 0) == 0,
    identified_pool_zero_assumed =
      is.na(identified_pool_tot_mean) & coalesce(identified_pool_observed_n, 0) == 0,

    grey_for_comparison = case_when(
      !is.na(grey_tot_mean) ~ grey_tot_mean,
      grey_zero_assumed ~ 0,
      TRUE ~ NA_real_
    ),
    harbour_for_comparison = case_when(
      !is.na(harbour_tot_mean) ~ harbour_tot_mean,
      harbour_zero_assumed ~ 0,
      TRUE ~ NA_real_
    ),
    identified_pool_for_comparison = case_when(
      !is.na(identified_pool_tot_mean) ~ identified_pool_tot_mean,
      identified_pool_zero_assumed ~ 0,
      TRUE ~ NA_real_
    ),

    species_sum_raised = grey_for_comparison + harbour_for_comparison,
    raised_difference = combined_tot_mean - species_sum_raised,
    percent_increase = case_when(
      is.finite(raised_difference) & species_sum_raised > 0 ~
        100 * raised_difference / species_sum_raised,
      TRUE ~ NA_real_
    ),
    pooling_effect = identified_pool_for_comparison - species_sum_raised,
    unresolved_increment = combined_tot_mean - identified_pool_for_comparison,
    comparison_available = is.finite(raised_difference)
  ) %>%
  arrange(desc(raised_difference))

fwrite(
  as.data.table(comparison_area_metier),
  "results/pinniped_case_raised_difference_area_metier.csv",
  sep = ";",
  na = "NA"
)

comparison_area_metier %>%
  select(
    ecoregion, areacode, metierl4,
    species_sum_raised, combined_tot_mean,
    raised_difference, percent_increase,
    pooling_effect, unresolved_increment,
    comparison_available
  )
```

Summaries retain métier, area, and grand-total views. Percentage increases are calculated from the corresponding summed estimates; percentages themselves are not summed.

```{r comparison-summaries}
safe_sum <- function(x) {
  if (all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
}

add_percent_increase <- function(x) {
  x %>%
    mutate(
      percent_increase = case_when(
        is.finite(raised_difference) & species_sum_raised > 0 ~
          100 * raised_difference / species_sum_raised,
        TRUE ~ NA_real_
      )
    )
}

comparison_by_metier <- comparison_area_metier %>%
  group_by(metierl4) %>%
  summarise(
    species_sum_raised = safe_sum(species_sum_raised),
    combined_raised = safe_sum(ifelse(comparison_available, combined_tot_mean, NA_real_)),
    combined_raised_all = safe_sum(combined_tot_mean),
    raised_difference = safe_sum(raised_difference),
    pooling_effect = safe_sum(pooling_effect),
    unresolved_increment = safe_sum(unresolved_increment),
    n_area_metier = n(),
    n_comparable = sum(comparison_available),
    .groups = "drop"
  ) %>%
  add_percent_increase() %>%
  arrange(desc(raised_difference))

comparison_by_area <- comparison_area_metier %>%
  group_by(ecoregion, areacode) %>%
  summarise(
    species_sum_raised = safe_sum(species_sum_raised),
    combined_raised = safe_sum(ifelse(comparison_available, combined_tot_mean, NA_real_)),
    combined_raised_all = safe_sum(combined_tot_mean),
    raised_difference = safe_sum(raised_difference),
    pooling_effect = safe_sum(pooling_effect),
    unresolved_increment = safe_sum(unresolved_increment),
    n_metiers = n(),
    n_comparable = sum(comparison_available),
    comparison_coverage = n_comparable / n_metiers,
    .groups = "drop"
  ) %>%
  add_percent_increase() %>%
  arrange(desc(raised_difference))

comparison_total <- comparison_area_metier %>%
  summarise(
    species_sum_raised = safe_sum(species_sum_raised),
    combined_raised = safe_sum(ifelse(comparison_available, combined_tot_mean, NA_real_)),
    combined_raised_all = safe_sum(combined_tot_mean),
    raised_difference = safe_sum(raised_difference),
    pooling_effect = safe_sum(pooling_effect),
    unresolved_increment = safe_sum(unresolved_increment),
    n_area_metier = n(),
    n_comparable = sum(comparison_available),
    comparison_coverage = n_comparable / n_area_metier
  ) %>%
  add_percent_increase()

comparison_by_metier
comparison_by_area
comparison_total

fwrite(
  as.data.table(comparison_by_metier),
  "results/pinniped_case_raised_difference_sum_over_areas_by_metier.csv",
  sep = ";", na = "NA"
)

fwrite(
  as.data.table(comparison_by_area),
  "results/pinniped_case_raised_difference_sum_over_metiers_by_area.csv",
  sep = ";", na = "NA"
)

fwrite(
  as.data.table(comparison_total),
  "results/pinniped_case_raised_difference_total.csv",
  sep = ";", na = "NA"
)
```

# 9. Visualise the comparison

The figures are driven by the current results rather than hard-coded areas or ecoregions.

## 9.1 Largest area x métier differences

```{r standout-area-metier}
standout_area_metier <- comparison_area_metier %>%
  filter(comparison_available, raised_difference > 0) %>%
  slice_max(
    order_by = raised_difference,
    n = N_STANDOUT_AREA_METIER,
    with_ties = FALSE
  ) %>%
  mutate(
    area_metier = paste(areacode, toupper(metierl4), sep = " | "),
    area_metier = forcats::fct_reorder(area_metier, raised_difference)
  )

if (nrow(standout_area_metier) > 0) {
  p_pinniped_case_standout_area_metier <- ggplot(
    standout_area_metier,
    aes(x = raised_difference, y = area_metier, fill = raised_difference)
  ) +
    geom_col(show.legend = FALSE) +
    scale_fill_gradientn(
      colours = c(ICES_SEASHELL, ICES_HORIZON, ICES_BLUE)
    ) +
    scale_x_continuous(
      labels = scales::label_number(accuracy = 1, big.mark = " "),
      expand = expansion(mult = c(0, 0.05))
    ) +
    labs(
      x = "Raised difference: combined - (grey + harbour)",
      y = "ICES area | métier level 4",
      title = "Largest area x métier differences",
      subtitle = paste0("Top ", nrow(standout_area_metier), " positive differences")
    ) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(face = "bold", colour = ICES_BLUE)
    )

  print(p_pinniped_case_standout_area_metier)

  ggsave(
    "results/pinniped_case_standout_raised_difference_area_metier.png",
    p_pinniped_case_standout_area_metier,
    width = 10,
    height = max(6, nrow(standout_area_metier) * 0.30 + 2),
    dpi = 300
  )
}
```

## 9.2 Raised estimates across cases

This uses the **same area x métier combinations as the ranked plot above**, so the two figures answer complementary questions: how large is the combined-minus-species difference, and what do the underlying raised estimates and uncertainty intervals look like?

By default only the three reporting cases are shown. Set `SHOW_DIAGNOSTIC_CASE <- TRUE` in the controls to add the identified-only pooled diagnostic. Cases without a BEAM raised estimate are not given an artificial interval and therefore do not appear.

```{r case-comparison-area-metier}
selected_area_metier <- standout_area_metier %>%
  distinct(ecoregion, areacode, metierl4)

case_plot_data <- pinniped_results %>%
  filter(
    total_status == "raised estimate available",
    is.finite(tot_mean),
    is.finite(tot_lwr),
    is.finite(tot_upr),
    SHOW_DIAGNOSTIC_CASE | case_role == "reporting"
  ) %>%
  semi_join(
    selected_area_metier,
    by = c("ecoregion", "areacode", "metierl4")
  ) %>%
  mutate(
    areacode = as.character(areacode),
    metierl4 = toupper(as.character(metierl4)),
    area_metier = paste(areacode, metierl4, sep = " | "),
    tot_lwr_plot = pmax(tot_lwr, 0)
  )

case_levels <- c(
  "Grey seal",
  "Harbour seal",
  "Grey + harbour (identified only; diagnostic)",
  "Grey + harbour + unresolved pinnipeds"
)

if (!SHOW_DIAGNOSTIC_CASE) {
  case_levels <- case_levels[
    case_levels != "Grey + harbour (identified only; diagnostic)"
  ]
}

if (nrow(case_plot_data) > 0) {
  # Match the row order used in the ranked-difference figure.
  area_levels <- standout_area_metier %>%
    arrange(raised_difference) %>%
    transmute(area_metier = paste(areacode, toupper(metierl4), sep = " | ")) %>%
    distinct() %>%
    pull(area_metier)

  # Small vertical offsets keep case intervals readable within each row.
  case_offsets <- seq(-0.24, 0.24, length.out = length(case_levels))
  names(case_offsets) <- case_levels

  case_plot_data <- case_plot_data %>%
    mutate(
      case_label = factor(case_label, levels = case_levels),
      area_metier = factor(area_metier, levels = area_levels),
      y_base = as.numeric(area_metier),
      y_position = y_base + case_offsets[as.character(case_label)]
    ) %>%
    filter(!is.na(case_label), !is.na(y_position))

  p_pinniped_case_case_comparison <- ggplot(
    case_plot_data,
    aes(colour = case_label)
  ) +
    geom_segment(
      aes(
        x = tot_lwr_plot,
        xend = tot_upr,
        y = y_position,
        yend = y_position
      ),
      linewidth = 0.6
    ) +
    geom_point(
      aes(x = tot_mean, y = y_position),
      size = 2.4
    ) +
    scale_x_sqrt(
      labels = scales::label_number(accuracy = 1, big.mark = " "),
      expand = expansion(mult = c(0, 0.05))
    ) +
    scale_y_continuous(
      breaks = seq_along(area_levels),
      labels = area_levels,
      expand = expansion(add = 0.6)
    ) +
    scale_colour_manual(
      values = CASE_COLOURS,
      drop = FALSE
    ) +
    labs(
      x = "Raised bycatch estimate (individuals; square-root scale)",
      y = "ICES area | métier level 4",
      title = "Raised estimates across pinniped cases",
      subtitle = paste0(
      "The" , length(area_levels),
        " higest ranked area x métier combinations; "
      ),
      colour = "Case"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(face = "bold", colour = ICES_BLUE)
    )

  print(p_pinniped_case_case_comparison)

  ggsave(
    "results/pinniped_case_raised_estimate_comparison_area_metier.png",
    p_pinniped_case_case_comparison,
    width = 11,
    height = max(6, length(area_levels) * 0.34 + 2.4),
    dpi = 300,
    limitsize = FALSE
  )
}
```

# 10. Maps

Two maps use the **same areas**:

1. absolute increase: `combined - (grey + harbour)`;
2. percentage increase relative to `grey + harbour`.

Area selection is controlled at the top of the walkthrough. `MAP_SHOW <- "top_n"` maps the `N_MAP_AREAS` largest positive absolute differences; `MAP_SHOW <- "all"` maps all positive differences. The percentage map does not re-rank the areas, so the two maps remain directly comparable.

The official ICES GIS area layer is queried directly because BEAM `areacode` can represent different ICES hierarchy levels. Each estimate is mapped at the finest level represented by its own code; larger-area estimates are not distributed into smaller polygons. If a future run contains ICES statistical-rectangle codes, the official ICES statistical-rectangle layer is added automatically.

The basemap and data fills use the ICES primary palette: a pale Horizon-blue sea, Seashell-white land, ICES-blue structure, and ICES-derived continuous scales for the estimates.

## 10.1 Prepare ICES geometry and select areas

```{r map-data}
normalise_ices_code <- function(x) {
  x <- tolower(trimws(as.character(x)))
  x[x == ""] <- NA_character_
  gsub("\\s+", "", x)
}

join_ices_code <- function(...) {
  parts <- list(...)
  n <- max(vapply(parts, length, integer(1)))
  parts <- lapply(parts, rep_len, length.out = n)

  vapply(seq_len(n), function(i) {
    z <- vapply(parts, `[`, character(1), i)
    z <- z[!is.na(z) & z != ""]
    if (length(z) == 0) NA_character_ else paste(z, collapse = ".")
  }, character(1))
}

read_ices_layer <- function(cache, url, fields) {
  refresh <- !file.exists(cache)

  if (!refresh) {
    cached <- try(sf::st_read(cache, quiet = TRUE), silent = TRUE)
    refresh <- inherits(cached, "try-error") || !all(fields %in% names(cached))
  }

  if (refresh) {
    message("Refreshing ICES spatial cache: ", basename(cache))
    response <- httr::GET(
      url,
      query = list(
        where = "1=1",
        outFields = paste(fields, collapse = ","),
        returnGeometry = "true",
        outSR = "4326",
        f = "geojson"
      )
    )
    httr::stop_for_status(response)
    writeBin(httr::content(response, as = "raw"), cache)
  }

  x <- sf::st_read(cache, quiet = TRUE) %>% sf::st_make_valid()
  missing_fields <- setdiff(fields, names(x))
  if (length(missing_fields) > 0) {
    stop("ICES layer is missing: ", paste(missing_fields, collapse = ", "))
  }
  x
}

make_lookup <- function(x, key, level, priority) {
  x %>%
    transmute(
      map_key = .data[[key]],
      spatial_level = level,
      match_priority = priority
    ) %>%
    filter(!is.na(map_key), map_key != "") %>%
    group_by(map_key, spatial_level, match_priority) %>%
    summarise(geometry = sf::st_union(geometry), .groups = "drop") %>%
    sf::st_as_sf()
}

# Candidate areas are ranked by the absolute increase.
map_area_candidates <- comparison_by_area %>%
  mutate(
    areacode = as.character(areacode),
    map_key = normalise_ices_code(areacode)
  ) %>%
  filter(
    is.finite(raised_difference),
    raised_difference > 0,
    !is.na(map_key)
  ) %>%
  arrange(desc(raised_difference))

if (nrow(map_area_candidates) == 0) {
  stop("No areas have a finite positive raised_difference.")
}

# Official ICES area hierarchy.
dir.create("data", recursive = TRUE, showWarnings = FALSE)
dir.create("results", recursive = TRUE, showWarnings = FALSE)

ices_area_fields <- c(
  "Major_FA", "SubArea", "Division", "SubDivisio", "Unit",
  "Area_Full", "Area_27"
)

ices_areas_sf <- read_ices_layer(
  cache = "data/pinniped_case_ices_areas.geojson",
  url = paste0(
    "https://gis.ices.dk/gis/rest/services/",
    "ICES_reference_layers/ICES_Areas/MapServer/0/query"
  ),
  fields = ices_area_fields
)

ices_areas_base <- ices_areas_sf %>%
  mutate(
    major_part = normalise_ices_code(Major_FA),
    subarea_part = normalise_ices_code(SubArea),
    division_part = normalise_ices_code(Division),
    subdivision_part = normalise_ices_code(SubDivisio),
    unit_part = normalise_ices_code(Unit),
    key_subarea = join_ices_code(major_part, subarea_part),
    key_division = join_ices_code(major_part, subarea_part, division_part),
    key_subdivision = join_ices_code(
      major_part, subarea_part, division_part, subdivision_part
    ),
    key_unit = join_ices_code(
      major_part, subarea_part, division_part, subdivision_part, unit_part
    ),
    key_area_full = normalise_ices_code(Area_Full),
    key_area_27 = {
      x <- normalise_ices_code(Area_27)
      ifelse(
        is.na(x),
        NA_character_,
        ifelse(grepl("^27\\.", x), x, paste0("27.", x))
      )
    }
  )

# Prefer an explicit hierarchy match over the generic fallback fields.
spatial_lookup_sf <- bind_rows(
  make_lookup(ices_areas_base, "key_unit", "ICES unit", 1),
  make_lookup(ices_areas_base, "key_subdivision", "ICES subdivision", 2),
  make_lookup(ices_areas_base, "key_division", "ICES division", 3),
  make_lookup(ices_areas_base, "key_subarea", "ICES subarea", 4),
  make_lookup(ices_areas_base, "key_area_full", "ICES full area", 5),
  make_lookup(ices_areas_base, "key_area_27", "ICES Area_27", 6)
)

# Add ICES statistical rectangles only when such codes occur in the results.
has_statrec_codes <- any(grepl("^[0-9]{2}[a-z][0-9]$", map_area_candidates$map_key))

if (has_statrec_codes) {
  ices_statrec_sf <- read_ices_layer(
    cache = "data/pinniped_case_ices_statrec.geojson",
    url = paste0(
      "https://gis.ices.dk/gis/rest/services/",
      "ICES_reference_layers/ICES_Statrec_Eco/MapServer/0/query"
    ),
    fields = c("ICESNAME", "Ecoregion")
  ) %>%
    mutate(statrec_key = normalise_ices_code(ICESNAME))

  spatial_lookup_sf <- bind_rows(
    make_lookup(ices_statrec_sf, "statrec_key", "ICES statistical rectangle", 0),
    spatial_lookup_sf
  )
}

# First identify what can be mapped, then apply the top-N choice. This avoids
# asking for 10 areas and silently receiving fewer because one did not join.
mappable_keys <- spatial_lookup_sf %>%
  sf::st_drop_geometry() %>%
  distinct(map_key)

# Wrong Trousers check: these area codes did not find the expected spatial fit.
unmatched_map_areacodes <- map_area_candidates %>%
  anti_join(mappable_keys, by = "map_key")

fwrite(
  as.data.table(unmatched_map_areacodes),
  "results/pinniped_case_unmatched_map_areacodes.csv",
  sep = ";",
  na = "NA"
)

map_area_candidates <- map_area_candidates %>%
  semi_join(mappable_keys, by = "map_key")

if (nrow(map_area_candidates) == 0) {
  stop("None of the positive area differences could be matched to ICES geometry.")
}

map_areas_to_show <- if (MAP_SHOW == "top_n") {
  map_area_candidates %>% slice_head(n = N_MAP_AREAS)
} else {
  map_area_candidates
}

# Join the selected estimates to geometry and keep the best hierarchy match.
map_results_sf <- spatial_lookup_sf %>%
  inner_join(map_areas_to_show, by = "map_key") %>%
  group_by(ecoregion, areacode, map_key) %>%
  filter(match_priority == min(match_priority)) %>%
  ungroup()

mapped_area_levels <- map_results_sf %>%
  sf::st_drop_geometry() %>%
  select(
    ecoregion, areacode, spatial_level,
    species_sum_raised, combined_raised,
    raised_difference, percent_increase,
    comparison_coverage
  ) %>%
  distinct() %>%
  arrange(desc(raised_difference))

fwrite(
  as.data.table(mapped_area_levels),
  "results/pinniped_case_mapped_area_levels.csv",
  sep = ";",
  na = "NA"
)

mapped_area_levels

# Background and map extent.
land_sf <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
map_labels_sf <- sf::st_point_on_surface(map_results_sf)
map_bbox <- sf::st_bbox(map_results_sf)

x_pad <- max(1, as.numeric(map_bbox["xmax"] - map_bbox["xmin"]) * 0.12)
y_pad <- max(1, as.numeric(map_bbox["ymax"] - map_bbox["ymin"]) * 0.12)

map_xlim <- c(map_bbox["xmin"] - x_pad, map_bbox["xmax"] + x_pad)
map_ylim <- c(map_bbox["ymin"] - y_pad, map_bbox["ymax"] + y_pad)

map_subtitle <- if (MAP_SHOW == "top_n") {
  paste0("Top ", nrow(map_results_sf), " mappable areas ranked by absolute increase")
} else {
  "All mappable areas with a positive absolute increase"
}
```

If `unmatched_map_areacodes` contains rows, they remain in the numerical comparison tables but are not shown on the map.

## 10.2 Absolute increase map

```{r map-absolute}
make_pinniped_map <- function(fill_var, title, legend_title, percent = FALSE, caption = NULL) {
  p <- ggplot() +
    theme_bw() +
    theme(
      panel.background = element_rect(
        fill = scales::alpha(ICES_HORIZON, 0.14),
        colour = NA
      ),
      panel.grid.major = element_line(colour = "white", linewidth = 0.25),
      panel.grid.minor = element_blank()
    ) +
    geom_sf(
      data = ices_areas_sf,
      fill = NA,
      colour = scales::alpha(ICES_GREEN, 0.55),
      linewidth = 0.15
    ) +
    geom_sf(
      data = map_results_sf,
      aes(fill = .data[[fill_var]]),
      colour = ICES_BLUE,
      linewidth = 0.55
    ) +
    geom_sf(
      data = land_sf,
      fill = ICES_SEASHELL,
      colour = scales::alpha(ICES_BLUE, 0.55),
      linewidth = 0.25
    ) +
    geom_sf_text(
      data = map_labels_sf,
      aes(label = areacode),
      size = 3,
      colour = ICES_BLUE,
      check_overlap = TRUE
    ) +
    coord_sf(xlim = map_xlim, ylim = map_ylim, expand = FALSE) +
    labs(
      title = title,
      subtitle = map_subtitle,
      caption = caption,
      x = NULL,
      y = NULL
    ) +
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", colour = ICES_BLUE),
      plot.subtitle = element_text(colour = "grey30"),
      plot.caption = element_text(hjust = 1)
    )

  if (percent) {
    p + scale_fill_gradient(
      low = ICES_BLUE,
      high = ICES_ORANGE,
      na.value = "grey90",
      name = legend_title,
      labels = scales::label_number(accuracy = 1, suffix = "%")
    )
  } else {
    p + scale_fill_gradient(
      low = ICES_BLUE,
      high = ICES_ORANGE,
      na.value = "grey90",
      name = legend_title,
      labels = scales::label_number(accuracy = 1, big.mark = " ")
    )
  }
}

p_pinniped_case_raised_difference_map <- make_pinniped_map(
  fill_var = "raised_difference",
  title = "Increase in the combined pinniped estimate",
  legend_title = "Combined -\n(grey + harbour)"
)

print(p_pinniped_case_raised_difference_map)

ggsave(
  "results/pinniped_case_raised_difference_map.png",
  p_pinniped_case_raised_difference_map,
  width = 10,
  height = 8,
  dpi = 300
)
```

## 10.3 Percentage increase map

This uses exactly the same selected areas as the absolute map. If the grey + harbour baseline is zero or non-finite, the percentage is undefined and is stored as `NA` rather than `Inf`/`NaN`.

```{r map-percentage}
if (any(is.finite(map_results_sf$percent_increase))) {
  p_pinniped_case_percent_increase_map <- make_pinniped_map(
    fill_var = "percent_increase",
    title = "Percentage increase in the combined pinniped estimate",
    legend_title = "Increase over\ngrey + harbour",
    percent = TRUE,
    caption = "Percentage is undefined where the grey + harbour baseline is zero or non-finite."
  )

  print(p_pinniped_case_percent_increase_map)

  ggsave(
    "results/pinniped_case_percent_increase_map.png",
    p_pinniped_case_percent_increase_map,
    width = 10,
    height = 8,
    dpi = 300
  )
} else {
  message("Percentage map skipped: no selected area has a finite percentage increase.")
}
```

# 11. Finish

Restore the standard BEAM object if the same R session will continue with other taxa.

```{r restore}
obs3 <- obs3_standard

if (USE_PARALLEL && exists("cl")) {
  parallel::stopCluster(cl)
}
```

Normal BEAM service is restored; the pinniped detour is complete.
