---
title: "BEAM_walkthrough_CommonDolphinBOBIC"
author: "WGBYC_ToRC"
date: "2026-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Common dolphin BOBIC ecoregion case study

```{r library load,  eval=FALSE}
# imports all R packages that are needed for BEAM
library(data.table) # data wrangling
library(reshape2) # data wrangling
library(icesConnect) # downloading data from ICES servers
library(httr) # parsing downloaded data above
library(jsonlite) # parsing downloaded data
library(glmmTMB) # model fitting
library(metafor) # testing heterogeneity 
library(ggeffects) # BPUE estimation
library(emmeans) # BPUE estimation, don't go above emmeans  #1.11.0 or it explodes!!!
library(this.path) # for auto-detecting the dir a script is executing out of
library(doSNOW) # parallel computation
library(doParallel) # parallel computation
library(progress) # lotion for impatient minds
library(this.path)
library(stringdist) # data checking and cleaning (string manipulation)
library(stringi)
#library(taxize) # data cleaning
library(taxizedb) # data cleaning, devtools::install_github("ropensci/taxizedb")
library(countrycode) # to convert between 2 and 3- letter country codes 
library(tcltk)
library(ggplot2)
library(reshape2)
library(systemfonts)
library(flextable) #not quite well built
library(scales)
library(countrycode)
library(stringi)
library(stringr)
library(dplyr)
library(tidyr)

```

```{r, load functions,  eval=FALSE}
setwd(this.path::this.dir())
source("lib/beam_lib.R")
```

```{r load credentials,  eval=FALSE}
ices_username <- readLines("username.txt")
ices_token<-readLines("token.txt")
```

```{r define years,  eval=FALSE}
years <- seq(2017, year(Sys.Date())-1)
```

```{r create cluster,  eval=FALSE}
detectCores()
ncores<-12 #change this to what you want

cl <- makeCluster(ncores)
registerDoSNOW(cl)
```

## Download the data
```{r get data, eval=FALSE}
all1 <- beam_get_raw_data("data/all1.csv", years = years, force_download = TRUE)
obs1 <- beam_get_raw_data("data/obs1.csv", years = years, force_download = TRUE)
bycatch1 <- beam_get_raw_data("data/bycatch1.csv", years = years, force_download = TRUE)
```

```{r 2025GBFix, eval=FALSE}

# Copy VL0010 rows from 2024
vl0010_2025 <- subset(
  all1,
  vesselLengthRange == "VL0010" & year == 2024 & country=="GB"
)
# Change year
vl0010_2025$year <- 2025
# Append to original data
all1 <- rbind(all1, vl0010_2025)


obs1$country[obs1$country=="GB-NIR"]<-"GB"

```


```{r 2025PTFix, eval=FALSE}

sel <- all1[
  ecoregion == "Bay of Biscay and the Iberian Coast" &
  metierL4 == "GNS" &
  country == "PT"
]

new <- rbind(
  copy(sel)[, `:=`(metierL4 = "GTR", daysAtSeaF = daysAtSeaF * 0.69)],
  copy(sel)[, daysAtSeaF := daysAtSeaF * 0.31]
)

all1 <- rbind(
  all1[!(ecoregion == "Bay of Biscay and the Iberian Coast" &
         metierL4 == "GNS" &
         country == "PT")],
  new
)

fwrite(all1,"data/all1_postcorrection.csv",sep = ";",na="NA")

```

## Data cleaning and preparation
```{r data clean, eval=FALSE}
source("lib/clean_data.R")
```

```{r merge bycatch and monitoring, eval=FALSE}
source("lib/generate_the_list.R")
```

```{r matching bycaught species and sampling protocol, eval=FALSE}
#code snippet from generate_the_list.R
obs3[,
    taxon_bycatch_monitor_ok := (taxa_monitored %in% c("all","elasmobranchs~seabirds~mammals", "protectedspecies")) | (taxa_monitored == taxon)
]

obs3[taxon == "elasmobranchs" & taxa_monitored == "fish",
     taxon_bycatch_monitor_ok := TRUE] #Taxon monitored is okay if fish was the protocol and they reported elasmobranchs

obs3[taxon == "fish" & taxa_monitored == "elasmobranchs~seabirds~mammals",
     taxon_bycatch_monitor_ok := FALSE] #Taxon monitored is not okay if fish were reported under elasmobranchs~seabirds~mammals

```

## BPUE estimation

```{r}
obs3_area_CD <- subset(obs3,(ecoregion=="bay of biscay and the iberian coast") &
                       (species=="delphinus delphis"))

area_CD_m4_spec <- unique(obs3_area_CD[, .(ecoregion, areacode, metierl4, species)])

bpue1_area_CD <- calc_bpue(needle = area_CD_m4_spec, 
                        cols = c("ecoregion", "areacode", "metierl4", "species"),
                        include.weights = T,
                        dat = obs3_area_CD[taxon_bycatch_monitor_ok==TRUE],
                        years = years)

fwrite(bpue1_area_CD, file = "data/bpue1_area_CD.csv", sep = ";",na="NA")

bpue2_area_CD <- annotate_bpue(bpue1_area_CD, cols = c("ecoregion", "areacode", "metierl4"))
fwrite(bpue2_area_CD, file = "data/bpue2_area_CD.csv", sep = ";",na="NA")
```

## Checking one of the models (gtr, 8a) by hand
Searching for other model candidates that may not contain sampling protocol, so we
can get an estimate

```{r}
dat <- subset(obs3_area_CD,metierl4=="gtr"& areacode %in% "27.8.a")
dat <- dat[daysatsea > 0]
dat$logDAS<-log(dat$daysatsea)
re_terms <- c("country", "areacode", "year", "quarter", "metierl5", 
              "vessellength_group", "samplingprotocol", "monitoringmethod")
re_terms <- re_terms[sapply(re_terms, function(x) length(unique(dat[[x]]))) >= 2]
re.i <- do.call(CJ, replicate(length(re_terms), c(TRUE, FALSE), simplify = FALSE)) 
re_candidates <- apply(re.i, 1, function(i) {
  if (all(i == FALSE)) return("1")
  terms <- sprintf("(1|%s)", re_terms[unlist(i)])
  return(paste("1 +", paste(terms, collapse = " + ")))
})							

candidates <- lapply(re_candidates, function(f_str) {
  curr_formula <- as.formula(sprintf("n_ind ~ %s", f_str))
  fit <- suppressMessages(tryCatch({
    m <- glmmTMB(formula = curr_formula, offset = logDAS, family = nbinom2, data = dat)
    
    
    vc <- VarCorr(m)$cond
    is_singular <- FALSE
    for (comp in vc) {
      if (any(diag(comp) < 1e-6)) is_singular <- TRUE # Variance near zero
      if (nrow(comp) > 1) {
        cor_mat <- attr(comp, "correlation")
        if (any(abs(cor_mat[lower.tri(cor_mat)]) > 0.99)) is_singular <- TRUE # Near perfect correlation
      }
    }
    
    if (!is_singular) return(m) else return("Singular") 
  }, error = function(e) e$message))
  return(fit)
})

valid_indices <- which(sapply(candidates, function(x) inherits(x, "glmmTMB")))


# Use AIC instead of BIC for more balanced complexity selection
valid_aics <- sapply(candidates[valid_indices], AIC)
valid_complexity <- lengths(regmatches(re_candidates[valid_indices], gregexpr("\\+", re_candidates[valid_indices])))

sel_table <- data.table(
  Index = valid_indices,
  AIC = valid_aics,
  Complexity = valid_complexity
)

sel_table <- sel_table[is.finite(AIC)]


min_aic <- min(sel_table$AIC)

# Zone of indifference (delta AIC < 2)
top_tier <- sel_table[AIC <= (min_aic + 2)]

setorder(top_tier, Complexity, AIC)
top_tier
top_tier$formula<-NA
for (i in 1:nrow(top_tier)){
  top_tier$formula[i]<-as.character(formula(candidates[[top_tier$Index[i]]]))[3]
}

#visualise by hand
top_tier 

# All the top tier models have the sampling protocol, so we do not produce estimates

```






## Total bycatch estimation, per yer and per quarter of the year
For this study case, we are using a modified `calc_total` function - coming from CIBBRINAs code (`calc_total_CD`), adapted for the species.

```{r}

source("lib/calc_total_CD.R")

#Manually creating a weights value here for compatibility with the CIBBRiNA code
obs3_area_CD[, weights := ifelse(year >= (max(years)-4),1,0.5)]

# Calculate totals for CD
years # Already loaded at the beggining of the walkthrough
quarters <- 1:4

comb <- expand.grid(quarter = quarters, year = years)

# modify colname for fishing effort
all_CD <- fread("data/fishing_all_years.csv")

all2_CD <- all_CD

colnames(all2_CD)[13] <- "daysatsea"

tot_CD <- NULL
for(i in 1:nrow(comb)){
  print(i)
  output <- calc_total_CD(bpue = bpue2_area_CD, 
                          analysis_resolution = c("ecoregion", "areacode", "metierl4", "species"),
                          dat = obs3_area_CD, # the same that was given to calc_bpue
                          fishing = all2_CD, 
                          #verbose = T,
                          #include.weights = T,
                          response = "n_ind",
                          effort_term = "daysatsea", 
                          #weights_values = "weights",
                          filter = list(year = comb$year[i], 
                                        quarter = comb$quarter[i]))
  tot_CD <- rbind(tot_CD, output)
}

fwrite(tot_CD, "results/tot_CD.csv")

```

## Reliability checks
Using WKMAMBY script, as it is an improved version 

#Reliability estimation (from WKMAMBY)
```{r}
#full_data <- fread("data/obs3_area_CD.csv")
full_data <- obs3_area_CD
#bpues_estimates <- fread("data/bpue1.csv")
bpues_estimates <- bpue2_area_CD

source("lib/reliability_estimation_wkmamby.R")
#This creates a bpues_estimates object

str(bpues_estimates)

#We now have the reliability of the BPUE based on the jackknife/rmse approach. We can add that to tot1, the same reliability outcome for each year of a given BPUE. 
#Then we compute the breadth of interval reliability on the tot estimate of each year, can combine them in an overall_reliability measure

tot_reliability <- tot_CD

tot_reliability[bpues_estimates,
                on = .(ecoregion, metierl4, species),
                reliability_rmse := i.reliability_rmse]

tot_reliability$tot_lwr_log <- log10(tot_reliability$tot_lwr+1)
tot_reliability$tot_upr_log<-log10(tot_reliability$tot_upr+1)
tot_reliability$delta<-tot_reliability$tot_upr_log-tot_reliability$tot_lwr_log

tot_reliability$reliability_CI <- tot_reliability$delta < 2

monitoring_cutoff <- bpue2_area_CD
monitoring_cutoff$monitoring_coverage <- monitoring_cutoff$daysatsea/monitoring_cutoff$daysatseaf

tot_reliability[monitoring_cutoff,
                on = .(ecoregion, metierl4, species),
                monitoring_coverage := i.monitoring_coverage]

tot_reliability$reliability_coverage <- tot_reliability$monitoring_coverage > 0.001

tot_reliability$overall_reliability <- tot_reliability$reliability_CI == TRUE & tot_reliability$reliability_rmse == TRUE & tot_reliability$reliability_coverage ==TRUE

#Need to bring in the zeros ?

tot_reliability[tot_mean == 0, overall_reliability := TRUE]

fwrite(tot_reliability, "results/tot_reliability_CD.csv")

```


## Quality control
Not done for this special case
```{r QC,eval=FALSE}
# tot2 <- tot_CD
# source("lib/QCs_TotalBycatch.R")
# source("lib/add_qc123.R")
# total_bycatch <- add_qc123(tot = tot_CD, 
#                            obs = obs3_area_CD, 
#                            all = all_CD, 
#                            cols = c("ecoregion", "year", "country", "metierl4", "metierl5"), 
#                            year_ass = (year(Sys.Date())-1))
# fwrite(total_bycatch, "data/final_beam_table_CD.csv", na="NA")

```


## Presenting and visualising results
```{r complete bycatch, eval=FALSE}

##here we create the risk table dynamically, accounting for the species list update and new monitoring

source("lib/risktable_making.R")
all.metier<-risktable_making(bycatch1,ecoreg_species)

all.metier$species<-tolower(all.metier$species)
all.metier$ecoregion<-tolower(all.metier$ecoregion)
THEDATA<-obs3_area_CD

tot_CD$complete<-"no"
tot_CD$missing<-0

whole.bycatch.df<-data.frame(ecoregion=tot_CD$ecoregion[!duplicated(tot_CD[,c(1,3)])],species=tot_CD$species[!duplicated(tot_CD[,c(1,3)])])

tot_CD.sub<-subset(tot_CD,!is.na(tot_mean))

for (i in 1:nrow(whole.bycatch.df)) {
  fished<-all.metier$metierL4[all.metier$ecoregion==whole.bycatch.df$ecoregion[i] &all.metier$species==whole.bycatch.df$species[i]] 
tb<-tot_CD.sub$metierl4[tot_CD.sub$ecoregion==whole.bycatch.df$ecoregion[i] &tot_CD.sub$species==whole.bycatch.df$species[i]] 
fished<-fished[!is.na(fished)]

if (all(fished%in%tb)==TRUE) {
tot_CD$complete[tot_CD$ecoregion==whole.bycatch.df$ecoregion[i] &tot_CD$species==whole.bycatch.df$species[i]] <-"yes"

}
tot_CD$missing[tot_CD$ecoregion==whole.bycatch.df$ecoregion[i] &tot_CD$species==whole.bycatch.df$species[i]] <-sum((fished%in%tb)==FALSE)


}

complete_tb<-tot_CD[!is.na(tot_CD$tot_mean)&tot_CD$complete=="yes",]

fwrite(complete_tb,"results/cases_with_complete_tot_CD.csv",na="NA")
fwrite(tot_CD,"results/cases_with_tot_CD.csv",na="NA")

```

