---
title: "BEAM walkthrough wgbyc 2026"
author: "David Lusseau"
date: "2026-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#load library

```{r library load,  eval=FALSE}
# imports all R packages that are needed for BEAM
library(data.table) # data wrangling
library(reshape2) # data wrangling
library(icesConnect) # downloading data from ICES servers
library(httr) # parsing downloaded data above
library(jsonlite) # parsing downloaded data
library(glmmTMB) # model fitting
library(metafor) # testing heterogeneity 
library(ggeffects) # BPUE estimation
library(emmeans) # BPUE estimation, don't go above emmeans  #1.11.0 or it explodes!!!
library(this.path) # for auto-detecting the dir a script is executing out of


library(doSNOW) # parallel computation
library(doParallel) # parallel computation
library(progress) # lotion for impatient minds
library(this.path)
library(stringdist) # data checking and cleaning (string manipulation)
library(stringi)
#library(taxize) # data cleaning
library(taxizedb) # data cleaning, devtools::install_github("ropensci/taxizedb")
library(countrycode) # to convert between 2 and 3- letter country codes 
library(tcltk)
library(ggplot2)
library(reshape2)
library(systemfonts)
library(flextable) #not quite well built
library(scales)
library(countrycode)
library(stringi)
library(stringr)
library(dplyr)
library(tidyr)

```

we then set the working directory and load the key functions

```{r, load functions,  eval=FALSE}
setwd(this.path::this.dir())
source("lib/beam_lib.R")

```


For regular implementation: once you have been granted access you will receive an email to login at https://data.ices.dk/token  
There you will generate a token which you need to copy and paste in the object below , eg, token<-"ewioubreuicvepb" which you can call in the function  
do note the username: it should be the email (eg louis.attack@dartmouth.edu) you used to make the access request  

```{r load credentials,  eval=FALSE}
ices_username <- readLines("username.txt")
ices_token<-readLines("token.txt")

```

we then set the range of years considered in the assessment. usually it is set as:
```{r define years,  eval=FALSE}
years <- seq(2017, year(Sys.Date())-1)
```


```{r create cluster,  eval=FALSE}
detectCores()
ncores<-10 #change this to what you want

cl <- makeCluster(ncores)
registerDoSNOW(cl)
```


###############################################################################################################
## Download the data
###############################################################################################################

```{r get data, eval=FALSE}
all1 <- beam_get_raw_data("data/all1.csv", years = years, force_download = TRUE)
obs1 <- beam_get_raw_data("data/obs1.csv", years = years, force_download = FALSE)
bycatch1 <- beam_get_raw_data("data/bycatch1.csv", years = years, force_download = FALSE)
```

```{r 2025GBFix, eval=FALSE}

# Copy VL0010 rows from 2024
vl0010_2025 <- subset(
  all1,
  vesselLengthRange == "VL0010" & year == 2024 & country=="GB"
)
# Change year
vl0010_2025$year <- 2025
# Append to original data
all1 <- rbind(all1, vl0010_2025)


obs1$country[obs1$country=="GB-NIR"]<-"GB"

```


```{r 2025PTFix, eval=FALSE}

sel <- all1[
  ecoregion == "Bay of Biscay and the Iberian Coast" &
  metierL4 == "GNS" &
  country == "PT"
]

new <- rbind(
  copy(sel)[, `:=`(metierL4 = "GTR", daysAtSeaF = daysAtSeaF * 0.69)],
  copy(sel)[, daysAtSeaF := daysAtSeaF * 0.31]
)

all1 <- rbind(
  all1[!(ecoregion == "Bay of Biscay and the Iberian Coast" &
         metierL4 == "GNS" &
         country == "PT")],
  new
)

fwrite(all1,"data/all1_postcorrection.csv",sep = ";",na="NA")



```

###############################################################################################################
## Data cleaning and preparation
###############################################################################################################


```{r data clean, eval=FALSE}
source("lib/clean_data.R")
```

we then generate the list of species from curated lists of species x ecoregion for ICES areas and the Mediterranean sea.it creates soecies_info.csv in the results folder which list species with their taxon and the ecoregion in which they are considered.  

```{r merge bycatch and monitoring, eval=FALSE}
source("lib/generate_the_list.R")

```


cases where taxon_bycatch_monitor_ok is FALSE will not be considered.  

###############################################################################################################
## BPUE estimation
###############################################################################################################

```{r bpue estimate,  eval=FALSE}

eco_m4_spec <- unique(obs3[, .(ecoregion, metierl4, species)])

bpue1 <- calc_bpue(eco_m4_spec,
                   cols = c("ecoregion", "metierl4", "species"),
                   dat = obs3[taxon_bycatch_monitor_ok==TRUE],
                   years=years,
                  include.weights = TRUE)
fwrite(bpue1, file = "data/bpue1.csv", sep = ";",na="NA")

```

########################################################################################
#########################################################################################

#### 2026 Baltic work ####


```{r areabase,eval=FALSE}
################################################################################
# For cases where ecoregion-level estimation and prediction is considered 
# inappropriate, we can go to the areacode level.

#### we want you to think about it for this step, so we kept it manual

#obs3_area <- subset(obs3,(ecoregion=="baltic sea") &
#                       (taxon=="mammals"))

obs3_area <- subset(obs3,(ecoregion=="baltic sea") &
                       (species =="pusa hispida" | species =="phocoena phocoena"))


area_m4_spec <- unique(obs3_area[, .(ecoregion, areacode, metierl4, species)])

bpue1_area <- calc_bpue(needle = area_m4_spec, 
                        cols = c("ecoregion", "areacode", "metierl4", "species"),
                        dat = obs3_area[taxon_bycatch_monitor_ok==TRUE])

fwrite(bpue1_area, file = "data/bpue1_area.csv", sep = ";",na="NA")

################################################################################

bpue2_area <- annotate_bpue(bpue1_area, cols = c("ecoregion", "areacode", "metierl4"))

#fwrite(bpue2_area, file = "data/bpue2_area.csv", sep = ";",na="NA")

```


```{r annotate bpue1, eval=FALSE}
bpue2 <- annotate_bpue(bpue1, cols = c("ecoregion", "metierl4"))
fwrite(bpue2, file = "data/bpue2.csv", sep = ";",na="NA")
```

##############################
### can we qualify monitored zero bycatch events?
##############################


```{r zero assessments, eval=FALSE}
#bpue1 <- fread("data/bpue1.csv")
#bpue2 <- fread("data/bpue2.csv")
source("lib/zeros_assessment.R")
zero_assessments(bpue2=bpue2,bpue1=bpue1)

```

###############################################################################################################
## Bycatch estimation
###############################################################################################################

we then attempt to predict the total bycatch that would result from the observed fishing effort that year given the estimated BPUE(s)  

```{r, eval=FALSE}
bpue1 <- fread("data/bpue1.csv")
```


```{r total bycatch, eval=FALSE}
tot1 <- vector(mode = "list", length = nrow(bpue1))

for (i in 1:nrow(bpue1)) {
    cat("\rProcessing row ", i, "/", nrow(bpue1), sep="")
    tot1[[i]] <- calc_total(bpue1[i],
                            obs = obs3[taxon_bycatch_monitor_ok==TRUE], 
                            all = all2, 
                            years = years,
                            verbose = FALSE)
}
tot1 <- rbindlist(tot1)
fwrite(tot1, file = "data/tot1.csv", sep = ";",na="NA")

```

#############################################################################



###############################################################################################################
## Quality control
###############################################################################################################

```{r QC,eval=FALSE}
tot2<-tot1
source("lib/QCs_TotalBycatch.R")
source("lib/add_qc123.R")
total_bycatch <- add_qc123(tot = tot1, obs = obs3, all = all2, cols = c("ecoregion", "year", "country", "metierl4", "metierl5"), year_ass = (year(Sys.Date())-1))
fwrite(total_bycatch, "data/final_beam_table.csv",na="NA")


```


note further work takes place in WGBYC ToR E to synthesis information about context for cases for which we can produce a BPUE or a Bycatch estimate or not.


###############################################################################################################
## Presenting and visualising results
###############################################################################################################


```{r complete bycatch, eval=FALSE}

##here we create the risk table dynamically, accounting for the species list update and new monitoring

source("lib/risktable_making.R")
all.metier<-risktable_making(bycatch1,ecoreg_species)
#all.metier<-fread("data/ecoregion_metier_sps_risk_bycatch_2023.csv")

#all.metier contains for each ecoregion x species a list of all metiers in which the species has been recorded as bycaught in any ecoregions and for which the metiers have fishing effort in the ecoregion of interest

all.metier$species<-tolower(all.metier$species)
all.metier$ecoregion<-tolower(all.metier$ecoregion)
THEDATA<-obs3

total_bycatch$complete<-"no"
total_bycatch$missing<-0

whole.bycatch.df<-data.frame(ecoregion=total_bycatch$ecoregion[!duplicated(total_bycatch[,c(1,3)])],species=total_bycatch$species[!duplicated(total_bycatch[,c(1,3)])])

total_bycatch.sub<-subset(total_bycatch,!is.na(tot_mean))

for (i in 1:nrow(whole.bycatch.df)) {

fished<-all.metier$metierL4[all.metier$ecoregion==whole.bycatch.df$ecoregion[i] &all.metier$species==whole.bycatch.df$species[i]] 
tb<-total_bycatch.sub$metierl4[total_bycatch.sub$ecoregion==whole.bycatch.df$ecoregion[i] &total_bycatch.sub$species==whole.bycatch.df$species[i]] 
fished<-fished[!is.na(fished)]

if (all(fished%in%tb)==TRUE) {
total_bycatch$complete[total_bycatch$ecoregion==whole.bycatch.df$ecoregion[i] &total_bycatch$species==whole.bycatch.df$species[i]] <-"yes"

}
total_bycatch$missing[total_bycatch$ecoregion==whole.bycatch.df$ecoregion[i] &total_bycatch$species==whole.bycatch.df$species[i]] <-sum((fished%in%tb)==FALSE)


}

complete_tb<-total_bycatch[!is.na(total_bycatch$tot_mean)&total_bycatch$complete=="yes",]

fwrite(complete_tb,"results/cases_with_complete_total_bycatch.csv",na="NA")
fwrite(total_bycatch,"results/cases_with_total_bycatch.csv",na="NA")


```

we then have a function to produce figures and tables of total estimates all saved in the results section:

```{r visualise summary, eval=FALSE}
endyear<-2025 #this is altered for the benchmark data

summaryyear<-aggregate(cbind(n_ind,daysatsea)~ecoregion+metierl4+species,sum,data=obs3[taxon_bycatch_monitor_ok==TRUE&year==endyear] )
summaryall<-aggregate(cbind(n_ind,daysatsea)~ecoregion+metierl4+species,sum,data=obs3[taxon_bycatch_monitor_ok==TRUE] )

names(summaryall)[4:5]<-c("n_ind_all","daysatsea_all")

#produce main table to print and bycatch estimate figures by taxon
source("lib/produce_table_and_overall_figures.R")

#postestimate_qa

```


###############################################################################################################
## Reliability measures - added September 2025
#######################################################################################################
```{r reliability,eval=FALSE}

library(glmmTMB)
library(ggeffects)
library(emmeans)
library(data.table)

##bpue_estimates
full_data <-obs3
bpues_estimates <-bpue1

source("lib/reliability_estimation_wkmamby.R")



tot_reliability <- tot1

tot_reliability[bpues_estimates,
                on = .(ecoregion, metierl4, species),
                reliability_rmse := i.reliability_rmse]

tot_reliability$tot_lwr_log <- log10(tot_reliability$tot_lwr+1)
tot_reliability$tot_upr_log<-log10(tot_reliability$tot_upr+1)
tot_reliability$delta<-tot_reliability$tot_upr_log-tot_reliability$tot_lwr_log

tot_reliability$reliability_CI <- tot_reliability$delta < 2

monitoring_cutoff <- bpue2
monitoring_cutoff$monitoring_coverage <- monitoring_cutoff$daysatsea/monitoring_cutoff$daysatseaf

tot_reliability[monitoring_cutoff,
                on = .(ecoregion, metierl4, species),
                monitoring_coverage := i.monitoring_coverage]

tot_reliability$reliability_coverage <- tot_reliability$monitoring_coverage > 0.001

tot_reliability$overall_reliability <- tot_reliability$reliability_CI == TRUE & tot_reliability$reliability_rmse == TRUE & tot_reliability$reliability_coverage ==TRUE

#Need to bring in the zeros ?

tot_reliability[tot_mean == 0, overall_reliability := TRUE]

fwrite(tot_reliability, "results/tot_reliability.csv")


```



#######################################################
#######################################################
#######################################################


```{r visualise, eval=FALSE}
library(tidyverse)
library(EnvStats)
library(ggalluvial)
library(colorspace)

#Prepares a new dataset for data visualisation, capturing failed checks along the way instead of discarding the data. Saves as new files in /data
source("lib/Data_viz_data_prep.R")
#Produces a series of figures showing effort suitability, availability of bycatch records, and ability to generate a BPUE across ecoregion, metier and taxon, saved in /results
source("lib/Data_viz.R")

#Interactive flow chart for sample retention through BEAM


library(shiny)
runApp("lib/Sample_retention_BEAM_flow_chart_2025.R")

```


all plots are saved in the results folder


this produces bpues_estimates saved as bpue_table_print_reliability_subset.csv in results folder  

We invite readers to read ICES WGBYC 2025 report for case studies which helps to understand this a bit more. We will aim to run through some of those case studies during the benchmark process.


we then merge this information back in the table which is then produced online as a html table (thanks to Carlos Pinto for producing it)

```{r production of final table for reporting, eval=FALSE}

temp<-merge(table.print,bpues_estimates,by.x=c("ecoregion","metierl4","species"),by.y=c("Ecoregion","metier L4","Species"),all.x=TRUE)

#We are now adding a color flag based on a bivariate scale combining information on reliability and monitoring coverage. The color column contain the HEX code for plotting, and the flag_description column indicates what the color means (e.g. uncertain estimate and low monitoring, more reliable estimate and monitoring above satisfactory theshold...)

# table.print$reliability<-FALSE
# table.print$reliability[which(temp$reliability2==TRUE)]<-TRUE
# 
# table.print <- add_color_flag(table.print, bpue2) #Uses monitoring coverage from bpue2 and reliability info present in table.print

#Convert TRUE/FALSE to qualifiers for printing
table.print$reliability <- ifelse(
  table.print$reliability,
  "more reliable",
  "uncertain"
)

#We are also flagging species of bycatch relevance - assessed in BEAM - that are known to be commerically landed in some ecoregions.

table.print <- add_landing_flag(table.print)

fwrite(table.print,file="results/bpue_table_print.csv",na="NA")

fwrite(subset(table.print,reliability=="more reliable"),file="results/table_reliable_estimates.csv",na="NA")


```



###############################################################################################################
## Reliability visualisation - added September 2025
###############################################################################################################


last we plot the total bycatch estimates for species x ecoregion x metier level 4 that have passed the reliability checks. we produce 3 types of plots:  
1. full plot with confidence bands in orange for the 'less reliable' estimates  
2. subset of only more reliable estimates  
3. subset of only more reliable estimates ordered by bycatch estimates (plotted below)  


